### 2017-05-28 paper/sexSelectionSpacing-ver1.tex
Move new Introduction into paper

% [ways to get at son preference:
% - stated fertility preferences
% - longer spacing after son
% - differential stopping behavior
% - biased sex ratio (overall before sex selection, and specifically at birth later)]
% 
% [why do these work less well after introduction of sex selection?]
% 
% [how can my method remedy those problems?]
% 
% [main results]

The effects of son preference has been an active research area for decades.
Before prenatal sex determination, a preference for sons was often associated 
with shorter birth spacing after the birth of a daughter than after the 
birth of a son, because that was the only way to achieve a son 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012}.
The resulting shorter spacing is associated with worse health outcomes 
for the girls 
\citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter, resulting in a lower parity progression
with the number of sons \citep{ben-porath76b,Das1987,Arnold1997,clark00}. 
Finally, increase in the ratio of males to females 
can serve as an indication of son preference because 
of differential mortality, for example because of the shorter
birth spacing, or underreporting \citep{sen90,sen92,Merli2000,anderson08}.

After the introduction of prenatal sex determination, a substantial
literature has focused on the use of prenatal sex selection as a direct 
measure of son preference.
Our understanding of sex selection usage is, however, constrained by a
lack of information;
there are no official data on sex selection, and the few surveys that ask 
about use of sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have instead used the sex of children 
born as the dependent variable, and estimate the effects of variables on the 
probability of having a son.
The idea is that in the absence of any interventions, the probability of having 
a son is approximately 0.512, and this is independent of genetic 
factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant coefficient therefore 
indicates an association between the variable and use of sex-selective 
abortions.
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

What has not previously been appreciated, however, is that the introduction 
of sex-selective abortions substantially changed the relationship between 
son preference and birth spacing.
The change happened because each abortion significantly increases the duration 
until the next birth.
This increase in duration can be divided into three parts starting from the time of
the abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Secondly, the average expected time to conception is about 6 months.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.%
\footnote{
There are three well-developed technologies for determining a fetus' sex: 
Chorionic villus sampling (CVS), amniocentesis, and ultrasound.
CVS can be applied after the shortest period
of gestation (8 to 12 weeks).
It is the most complicated, but also the most reliable, and an abortion can be done in 
the first trimester.
Amniocentesis can be performed after fourteen weeks, but requires three to four weeks
before the result is available, so an abortion cannot be performed until more than 
halfway through the second trimester.
Ultrasound has the advantages of being non-invasive and relatively cheap but 
is associated with higher risk of faulty sex determination if done early.
Generally, a fetus' sex can be determined in the third month of gestation if it is a
boy and the fourth month if it a girl.
}
The result is that each abortion delays the next birth by approximately 
a year.%
\footnote{
The waiting time to conception does vary by woman, but
even if it is very short, say one  month, the minimum additional space between
births would be 6 months per abortion.
For a discussion of time to conception see \citep{Wang2003}
}

% [why this matters]
How spacing interacts with sex selection is important for two reasons.
First, spacing, by itself, can no longer be used to predict son preference. 
Because each abortion extends spacing by about a year, we now have a situation 
where the \emph{longest} spacing is likely observed for families with the 
strongest son preference.
To further complicate matters, 
we may still observe short spacing after the birth of daughter as a representation 
of son preference for families who either have less access to sex-selective 
abortions or are willing to have more children.
Spacing is still important and useful in understanding sex selection and son
preference, but only if combined with the outcome of that spacing---in other words 
the likelihood of observing a boy or a girl.

% using spacing itself to predict son preference becomes useless if
% families have low desired fertility and access to sex-selective abortions!!!!
Second, if the duration from last birth becomes long enough, parents may 
reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
Hence, for a given parity, the sample of women for whom we do observe a birth 
may behave differently from the sample of women who have not yet had a birth.
This is a problem because births to women with shorter spacing are more likely 
to be in the survey for a given parity.
Hence, if decisions on sex selection change with duration since last birth,
the predicted sex ratio for a parity using the current method will be a 
biased estimate of the final observed sex ratio for the parity when 
childbearing is complete.

A number of factors make a reversal of the decision to use sex 
selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that a 
child of any sex is better than not having a child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although the latter two factor are most often associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

% [contribution of paper]
% [TK maybe not focus so much on the simple method but rather understanding son preference].
My main contribution here is the introduction and application
of a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the 
duration between births.
[TK move to estimation strategy?]
The empirical model is a competing risk, non-proportional hazard model with 
two exit states: either a boy or a girl is born.
This approach has TK major advantages over the simple method

Secondly, by explicitly incorporating censoring of birth spacing, it addresses any 
potential bias from ignoring how sex selection decisions may change with duration from 
the previous birth.

The method can better predict ``final'' sex ratios than the simple model

None of these is possible using the simple method.

>>>>

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for
parents who wanted a son, but did not yet have one, was to have the 
next birth sooner.
The result was shorter average spacing after the births of girls than boys,
and worse health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination substantially changed the relationship between 
son preference and birth spacing.
The issue is that each sex-selective abortion increases the duration between 
observed births by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}
We therefore now have a situation where families with the \emph{strongest} son 
preference likely show the \emph{longest} spacing after the birth of a daughter.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.
In this paper, I introduce and apply a novel method that
directly incorporates the effects of sex-selective abortions on
the likelihood of a son being born \emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.

How spacing interacts with sex selection is important for two reasons.
First, spacing, by itself, can no longer be used to measure son preference. 
Spacing is still important and useful in understanding son preference, but 
only if combined with the likelihood of observing a boy or a girl.
Short spacing after a girl is indicative of son preference if observed
births are close to the natural sex ratio, whereas long spacing is indicative
of son preference if observed births are male biased.

Second, the duration since last birth may affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, parents 
may reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
Hence, by examining whether sex selection decisions change with
spacing, we can draw a more nuanced picture of the degree of son preference.

% [contribution of paper]
% [TK maybe not focus so much on the simple method but rather understanding son preference].
My main contribution here is the introduction and application
of a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the 
duration between births.
[TK move to estimation strategy?]
The empirical model is a competing risk, non-proportional hazard model with 
two exit states: either a boy or a girl is born.
